---
# execute:
  # eval: false
---

Supervised Learning
===================

Introduction
------------

**Supervised learning** methods learn from training data in order to make
predictions about new data.

Before going further, let's review some necessary vocabulary:

* A **unit** is an entity of interest in a data set. Units are sometimes also
  called subjects.
* An **observation** is all measurements for a single unit. Observations are
  sometimes also called cases.
* A **feature** is a single measurement across all units. Features are
  sometimes also called variables, covariates, or predictors.
* A **response** is a feature that you want to predict.

For example, suppose you've collected a data set with the age, height,
and species of every tree in your neighborhood. Each tree is a unit. Each (age,
height, species) measurement for a tree is an observation. The features are
age, height, and species. If you want to use supervised learning to predict
height based on age and species, then height is a response. 🌲🌳🌴

Problems and methods are conventionally described in terms of the type of
response:

* **Regression** refers to a numerical response.
* **Classification** refers to a categorical response. We call the categories
  **classes**.

Continuing the trees example, if you want to predict height in meters, you have
a regression problem. If you want to predict species, you have a classification
problem.

This chapter focuses on classification, but many of the concepts covered here
are equally relevant to regression. Moreover, there are regression variants of
many classification methods. In fact, many classifiers are just a combination
of a regression method that predicts class probabilities and a decision rule
that selects a class based on the prediction.

::: {.callout-note}
If you want to learn more about regression, see DataLab's [Introduction to
Regression Modeling in R workshop series][dl-regression].

[dl-regression]: https://ucdavisdatalab.github.io/workshop_regression/
:::

::: {.callout-caution}
Classification methods are reductive: they assume that classes are mutually
exclusive and only predict one class for each data point.

In contexts where predictions are meant to inform people making decisions,
regression methods are often more appropriate and yield more informative
predictions (for example, class probabilities). Frank Harrell (Professor of
Biostatistics, Vanderbilt University) discusses this in his post
[Classification vs. Prediction][harrell].

[harrell]: https://www.fharrell.com/post/classification/

On the other hand, classification methods often work well for automating
decisions or tasks that are straightforward but tedious, especially if the
stakes are low.
:::

A supervised learning method consists of:

* A **model** for making predictions, with assumptions and adjustable
  **parameters** that govern what predictions are made.
* A **loss function** for measuring the real or abstract cost of incorrect
  predictions.

Methods "learn" by selecting parameter values that minimize loss for
predictions about training data, in the hope that these values will generalize
well to new data.

Most models also have **hyperparameters** (or tuning parameters), parameters
that are not learned and must instead be specified before training. Usually,
these control how much the model adapts to the training data. To select
hyperparameter values, use your own knowledge about the data as well as model
evaluation techniques (see @sec-evaluating-models).

The next section presents an overview of packages for supervised learning in R,
some of which are demonstrated in later examples. Subsequent sections explain
how to select features, how to select a model, and common model evaluation
strategies, including data partitioning and cross-validation. The chapter ends
with a short discussion of how to improve model performance and where to go to
learn more.


Packages
--------

The R community has developed dozens of machine learning packages to supplement
R's small selection of built-in functions. A majority of these provide a single
method or family of methods (such as decision trees).

The [CRAN Task View: Machine Learning page][cran-ml] is a good starting point
for finding packages related to supervised learning. The task view is regularly
updated by the CRAN administrators and only lists popular, actively-maintained
packages.

<!-- (for unsupervised learning, see TODO)-->

[cran-ml]: https://cran.r-project.org/view=MachineLearning

::: {.callout-tip}
If the task view doesn't list a package for a method you want to use, try
searching the web. Many packages hosted on CRAN or GitHub are not listed in the
task view because they're obscure, unmaintained, or in early development. None
of these things mean a package won't work, although you might want to check the
package code carefully to make sure it does what you expect.
:::

Different packages generally provide different programming interfaces. This
makes studying machine learning in R difficult: in addition to learning each
method, you must also learn the idiosyncrasies of each package. This also makes
it difficult to compare methods by swapping them in and out of a data
processing pipeline. To address this, some community members have developed
packages that provide a common programming interface on top of existing,
independent methods packages. Others have developed omnibus packages that
implement a wide variety of methods. These packages are listed below from most
to least CRAN downloads in March 2024.

* [caret][] is a common interface for training models with existing packages,
  as well as a collection of functions for related tasks (such as splitting
  data).
* [tidymodels][] is a metapackage (a collection of packages) for supervised
  learning designed to work well with [tidyverse][] packages. In some respects,
  this is the successor to caret. Especially important are the packages:
    * [parsnip][] provides a common interface for existing methods packages, or
      "engines" in the terminology of parsnip. For some methods, multiple
      engines are available, while for others, there is only one.
    * [rsample][] provides functions to partition and sample data sets.
    * [yardstick][] provides functions to compute a variety of performance
      statistics.
    * [broom][] provides functions to summarize model results in data frames.
* [h2o][] is an omnibus machine learning package, with implementations of many
  popular supervised and unsupervised learning methods. h2o is available for
  many different programming languages and places emphasis on computational
  efficiency.
* [mlr3][] and [mlr3learners][] provide a common, object-oriented interface for
  existing methods packages.
* [SuperLearner][] is an omnibus machine learning package designed to evaluate
  multiple methods and select the best combination of them.
* [qeML][] is a relatively new package that provides a common interface for
  existing methods, intended to be easier to use than tidymodels and mlr3. qeML
  is developed by [UC Davis Professor Emeritus Norm Matloff][norm].

[tidymodels]: https://www.tidymodels.org/
[tidyverse]: https://www.tidyverse.org/
[broom]: https://broom.tidymodels.org/
[parsnip]: https://parsnip.tidymodels.org/
[rsample]: https://rsample.tidymodels.org/
[yardstick]: https://yardstick.tidymodels.org/

[mlr3learners]: https://mlr3learners.mlr-org.com/
[mlr3]: https://mlr3.mlr-org.com/

[caret]: https://topepo.github.io/caret/
[SuperLearner]: https://cran.r-project.org/web/packages/SuperLearner/
[h2o]: https://cran.r-project.org/web/packages/h2o/
[qeML]: https://cran.r-project.org/web/packages/qeML/
[norm]: https://faculty.engineering.ucdavis.edu/matloff/

The caret, tidymodels, and h2o packages are all relatively complete solutions
with detailed documentation, many users, and many contributors.


### Example: Classifying Penguins {#sec-penguins}

![
  Chinstrap, Gentoo, and Adélie penguins. Artwork by [Allison Horst][horst].
](../images/lter_penguins.png)

[horst]: https://allisonhorst.com/

In this example, we'll fit a $k$-nearest neighbors (kNN) classifier with the
tidymodels package. In $k$-nearest neighbors, new observations are classified
by taking a majority vote of the classes of the $k$ nearest training set
observations. Generally, we choose odd $k$ to prevent ties, although there are
schemes for breaking ties if $k$ is even.

As an example data set, we'll use the Palmer Penguins data set, which was
collected by [Dr. Kristen Gorman][gorman] at [Palmer Station,
Antarctica][palmer]. The data set records physical characteristics for hundreds
of individual penguins from three different species: [Adélie][], [Chinstrap][],
and [Gentoo][].

[gorman]: https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php
[palmer]: https://pallter.marine.rutgers.edu/
[palmerpenguins]: https://allisonhorst.github.io/palmerpenguins/
[Adélie]: https://en.wikipedia.org/wiki/Ad%C3%A9lie_penguin
[Chinstrap]: https://en.wikipedia.org/wiki/Chinstrap_penguin
[Gentoo]: https://en.wikipedia.org/wiki/Gentoo_penguin

The data set is available for R in the [palmerpenguins][] package. We'll also
use several other packages in this example. To install them all:

```{r}
#| label: ex01_install
#| eval: false
install.packages("palmerpenguins")
install.packages("ggplot2")
install.packages("tidymodels")
```

Load the palmerpenguins package, which will automatically create a `penguins`
variable, and take a look at the first few observations:

```{r}
#| label: ex01_penguins
library("palmerpenguins")

head(penguins)
```

It's a good idea to do some exploratory analysis before fitting a model. We can
use ggplot2 to plot bill length, body mass, and species for each penguin:

```{r}
#| label: ex01_plot
library("ggplot2")

plt = ggplot(penguins) + geom_point() + aes(shape = species, color = species)

plt %+% aes(x = bill_length_mm, y = body_mass_g)

# Other plots that might be of interest:
# plt %+% aes(x = bill_length_mm, y = bill_depth_mm)
# plt %+% aes(x = bill_length_mm, y = flipper_length_mm)
# plt %+% aes(x = bill_depth_mm, y = body_mass_g)
# plt %+% aes(x = bill_depth_mm, y = flipper_length_mm)
# plt %+% aes(x = flipper_length_mm, y = body_mass_g)
```

The plot shows that the three species of penguins are relatively well-separated
by bill length and body mass. Adélie penguins have relatively short bills and
light bodies, Chinstrap penguins have relatively long bills and light bodies,
and Gentoo penguins have relatively long bills and heavy bodies. The species
only overlap for a few penguins. This separation is a sign that a classifier is
likely to work well.

:::{.callout-tip}
Classification is more difficult when the classes are not well-separated by the
features, but not necessarily intractible. Sometimes it's possible to transform
features to provide better separation. Some models can also take advantage of
interactions between features that are not readily apparent in scatter plots.
:::

Initial exploratory analysis is also when you should check data for outliers
and missing values. The previous plot didn't show any signs of extreme
outliers. Now let's check for missing values:

```{r}
#| label: ex01_check_na
summary(penguins)
```

kNN classifiers can't handle missing values, and only a few observation have
them, so let's remove these observations from the data set:

```{r}
#| label: ex01_remove_na

peng = na.omit(penguins)
```

Now let's train a kNN classifier with tidymodels. Specifically, we'll use the
parsnip package, which is included with tidymodels and provides a common
interface for a wide variety of supervised learning models.

Training with parsnip follows the same two steps for every model:

```{r}
#| label: ex01_train_parsnip
library("parsnip")

knn = nearest_neighbor("classification", neighbors = 10)        # <1>
fitted = fit(knn, species ~ bill_length_mm + body_mass_g, peng) # <2>
```
1. Initialize a model and set hyperparameters by calling a model function. In
   this case, we use `nearest_neighbor` since we want a kNN model, and set the
   number of neighbors to 10.
2. Train the model by calling `fit` with the model, a formula `~` that
   specifies the response on the left-hand side and the predictors on the
   right-hand side, and the data.


You can print the fitted model to get more information:

```{r}
#| label: ex01_fitted
fitted
```

This shows the type of response, an error rate estimate on the training set,
**kernel** ( distance-weighting function used in the class vote), and choice of
$k$ (misleadingly listed as "Best k"). The kernel is actually another
hyperparameter for this model, but parsnip selects a reasonable kernel by
default. You can find the hyperparameters and their default values for a model
on the model function's help page. In this case, that's `?nearest_neighbor`.

Once you've trained a model, you can use it with the `predict` function to make
predictions. The `predict` function takes the fitted model and a data set with
the same columns (excluding the response) as the training set as arguments. The
function returns the predictions in a 1-column data frame.

To try out `predict`, let's calculate the error on the training set manually.
First, predict the class for each observation in the training set:

```{r}
#| label: ex01_predict
preds = predict(fitted, peng)

preds
```

Notice that `predict` returns a data frame with a single column named
`.pred_class`. The return value is always a data frame with a `.pred_class`
column, regardless of the model used.

Now compute the proportion of incorrect predictions:

```{r}
#| label: ex01_error
sum(peng$species != preds$.pred_class) / nrow(peng)
```

Since the model was trained on these penguins, this error rate likely
underestimates the error rate for new penguins. Moreover, this is an overall
error rate and doesn't tell us much about error rates for the individual
species.

We can get a slightly better idea of the model's behavior by plotting the data
again, marking the misclassified penguins:

```{r}
#| label: ex01_error_plot
incorrect = peng$species != preds$.pred_class

ggplot(peng) + geom_point() +
  aes(x = bill_length_mm, y = body_mass_g, shape = species,
    color = incorrect) +
  scale_color_manual(values = c("gray", "red"))
```

The plot shows that most of the misclassified penguins are in regions where one
group overlaps with another. We'll revisit this data set and model in subsequet
examples, to learn about better evaluation methods and about strategies for
improving the fit.


Selecting a Model
-----------------

![
  Classification boundaries for ten different models across three different
  data sets. Visualization by [G. Varoquaux, A. Müller, & J.
  Grobler][sklearn-models].
](../figures/scikit_learn_classifiers.png)

[sklearn-models]: https://scikit-learn.org/stable/auto_examples/classification/plot_classifier_comparison.html#sphx-glr-auto-examples-classification-plot-classifier-comparison-py

The **No Free Lunch Theorem** states that there is no single "best" supervised
learning model. Different models make different assumptions, and will work
better for data that satisfy those assumptions. Generally, you should select a
model by comparing several different models and using whatever knowledge you
have about the data.

:::{.callout-caution}
The wide variety of models people have invented can be distracting. For
improving predictive performance, it's almost always more effective to collect
more data or to select or engineer better features than it is to switch models.
:::

In spite of the No Free Lunch Theorem, some models *are* more general than
others, and make good starting points. Here's a list of some well-known
classification models, where ✨ indicates a good starting point model:

* ✨ Decision trees ([classification tree guide][mlu-tree], [random forest
  guide][mlu-forest])
    * Pros: easy to interpret, can handle missing data, selects features
      automatically
    * Cons: prone to overfitting, sensitive to class imbalance, training is
      computationally expensive
    * Takeaway: random forests and boosted trees, which address overfitting at
      the cost of interpretability, perform well on a wide variety of problems
* ✨ Generalized linear models (GLM) ([logistic regression guide][mlu-logistic])
    * Pros: easy to interpret, training is computationally cheap,
      well-understood statistical properties, robust on small data sets
    * Cons: moderate assumptions (linearity)
    * Takeaway: GLMs have a long history but remain effective, especially for
      smaller data sets
* ✨ $k$-nearest neighbors (kNN)
    * Pros: conceptually simple
    * Cons: prediction is computationally expensive, poor performance at high
      number of features
    * Takeaway: appropriate for small to medium data sets
* Naive Bayes
    * Pros: conceptually simple, scales well
    * Cons: strong assumptions (independence), only for classification
    * Takeaway: useful for very large data sets
* Support vector machines (SVM)
    * Pros: okay with high number of features
    * Cons: hard to interpret, conceptually complex, binary by default
    * Takeaway: SVMs have become less popular than tree models in recent years,
      but can still be effective for many problems
* Discriminant analysis (DA)
    * Pros: well-understood statistical properties
    * Cons: strong assumptions (normality), only for classification
    * Takeaway: there are better models

[mlu-tree]: https://mlu-explain.github.io/decision-tree/
[mlu-forest]: https://mlu-explain.github.io/random-forest/
[mlu-logistic]: https://mlu-explain.github.io/logistic-regression/

:::{.callout-tip}
For another perspective on selecting a model, see [the scikit-learn algorithm
cheat sheet][sklearn-cheat].

Although scikit-learn is a Python package, [its user guide][sklearn-guide] is
also an excellent, concise reference for machine learning concepts (without any
code).
:::

[sklearn-cheat]: https://scikit-learn.org/stable/tutorial/machine_learning_map/
[sklearn-guide]: https://scikit-learn.org/stable/user_guide.html

For most models, there's a tradeoff between simplicity (in the sense of fewer
parameters) and flexibility.

Simple models make strong assumptions about the data, so there are fewer
parameters to estimate. If the assumptions are satisfied, simple models perform
well even when the number of observations is low or the observations are
relatively noisy. However, if the assumptions aren't satisfied, simple models
tend to **underfit**, meaning they fail to learn the signal in the training
data. Error in an underfit model is mostly **bias** caused by the assumptions.

:::{.callout-important}
Bias is a specific statistical term which doesn't carry the same connotation of
unfairness as the common English meaning. Instead, it means that on average, a
model's predicted values differ from the actual values.

Whether or not statistical bias makes a model unfair depends on the context in
which the model will be used. For contexts where it doesn't, there are often
good reasons to accept some bias.
:::

Flexible models make only weak assumptions about the data. If the number of
observations is high and the observations are not too noisy, flexible models
typically perform well. However, if the number of observations is low or the
observations are very noisy, flexible models tend to **overfit**, learning the
noise in the training data instead of the signal. As a result, the model will
not generalize well to new observations. Error in an overfit model is mostly
due to **variance**, meaning the model is too sensitive to small changes in the
training set.

Simple versus flexible is spectrum, and models usually provide hyperparameters
to control how simple or flexible the model is. One example is $k$ in
$k$-nearest neighbors. For $k$ close to the number of observations, the model
is simple and biased in favor of the majority class. For $k$ close to 1, the
model is flexible: it uses only a few observations in the training set to make
each prediction, and thus is susceptible to noisy or anomalous observations.

The tradeoff between simple and flexible models is more commonly known as the
**bias-variance tradeoff**. You can learn more about the bias-variance tradeoff
from [this interactive demo][mlu-bias-var].

[mlu-bias-var]: https://mlu-explain.github.io/bias-variance/


### Example: Bias-Variance Tradeoff

Simulations are a great way to answer questions about or demonstrate properties
of machine learning models. Let's use a simulation to demonstrate the
bias-variance tradeoff.

First, let's generate some data along this curve:

$$
y = x^3 - 10 e^{-0.5 (x - 3)^2} + 1
$$

Any non-polynomial curve will do, but we'll use this one because it plots
nicely. To generate points for $x$ in $[-1, 1]$ run:

```{r}
#| label: ex02_xy
n = 80
x = runif(n, -1, 1)
y = x^3 - 10 * exp(-0.5 * (x - 3)^2) + 1
```

Now let's make several copies of these points, with random noise added to the
$y$ coordinates, to simulate collecting multiple samples with measurement
error. To simulate 20 samples:

```{r}
#| label: ex02_data
n_samples = 20

x = rep(x, n_samples)

noise = rnorm(n * n_samples, 0, 0.05)
y = rep(y, n_samples) + noise

# Store the points in a data frame with sample ID.
sample_id = rep(seq(n_samples), each = n)
data = data.frame(x = x, y = y, sample_id = sample_id)
```

Before we fit any models, let's plot one of the samples to see what it looks
like:

```{r}
#| label: ex02_data_plot
sample01 = subset(data, sample_id == 1)

ggplot(sample01) + aes(x = x, y = y) + geom_point()
```

Now let's fit two models to each sample:

1. Since the data look somewhat quadratic, we'll fit a linear model with $x$
   and $x^2$ as features. This will be our simple, high-bias model.
2. We'll also fit a model that's just the mean of every 3 points. Although this
   model is conceptually simple, it will be our flexible, high-variance model.

We'll use the following function to fit the models. Note that it uses the zoo
package's `rollmean` function to fit the 3-point mean model, where "fit" really
just means computing the means. For the linear model, the function fits the
model with `lm` and then computes points on the regression line for a grid of
$x$ coordinates.

```{r}
#| label: ex02_fit_models_func
# install.packages("zoo")
library("zoo")

fit_models = function(samp) {
  # Sort the points by x coordinate.
  samp = samp[order(samp$x), ]

  # Fit the linear model.
  fitted_lm = lm(y ~ 1 + x + I(x^2), samp)

  # Evaluate the linear model on a grid.
  grid = seq(min(samp$x), max(samp$x), length.out = 1000)
  lm_pts = data.frame(x = grid)
  lm_pts$yhat = predict(fitted_lm, lm_pts)
  lm_pts$model = "linear"

  # Fit the 3-point mean model.
  mean3_pts = data.frame(
    # All x except first and last.
    x = samp$x[-c(1, nrow(samp))],
    yhat = rollmean(samp$y, 3),
    model = "3-point mean"
  )

  # Combine into a single data frame.
  fitted = rbind(lm_pts, mean3_pts)
  fitted$sample_id = samp$sample_id[[1]]
  fitted
}
```

We can us the `fit_models` with `split` and `lapply` to fit the models to each
sample, and then combine all of the predictions into a single data frame:

```{r}
#| label: ex02_fit_models
by_sample = split(data, data$sample_id)

fitted = lapply(by_sample, fit_models)
fitted = do.call(rbind, fitted)
```

Finally, let's plot the fitted models:

```{r}
#| label: ex02_fit_plot
ggplot(fitted) +
  # Plot the points from one sample as an example.
  geom_point(aes(x = x, y = y), sample01, color = "gray") +
  # Plot the fitted models.
  geom_line(aes(x = x, y = yhat, group = sample_id, color = factor(sample_id)),
    alpha = 0.5) +
  facet_wrap(vars(model)) +
  guides(color = "none")
```

The plot shows that the 3-sample mean model varies far more across samples than
the linear model. In other words, the 3-sample mean model has high variance.
The linear model is more stable, but doesn't follow the training data as
well---it has high bias.

Simulations like this one are an important tool in every machine learning
practicioner's toolkit. You can use them to get a rough idea of how a model or
estimator will behave under specific conditions. This example is based on one
from [Feature Engineering and Selection: A Practical Approach for Predictive
Models][fes] by M. Kuhn & K. Johnson.

[fes]: https://bookdown.org/max/FES/



### Example: Four Penguin Models {#sec-four-models}

Let's revisit the Palmer penguins data from @sec-penguins. This time, we'll fit
several different models, in order to examine how much accuracy differs between
them.

In order to get reliable accuracy estimates, let's randomly split the data into
two parts: a training set and a test set. The training set will contain 75% of
the observations and the test set will contain the remaining 25%.  We'll train
the models on the training set, and estimate accuracy on the test set. This
ensures that we won't overestimate the accuracy.

Splitting data into a training set and test set is a common practice. The
rsample package, which is included in tidymodels, provides an `initial_split`
function to split a data set. The default split is 75-25. You can use the
accompanying `training` and `testing` functions to get the training set and
test set from the split object:

```{r}
#| label: ex03_split
library("rsample")

# Set a seed to make the split reproducible.
set.seed(10332)
splitted = initial_split(peng)
peng_train = training(splitted)
peng_test = testing(splitted)
```

Now let's fit the models and compute error estimates. This time, we'll use all
four numeric features in the data set: bill length, bill depth, flipper length,
and body mass. We'll use the default hyperparameters for each model except
$k$-nearest neighbors, where there is no default for $k$. We'll use $k = 15$,
since a rule of thumb for choosing $k$ is to use $\sqrt{n}$, where $n$ is the
number of observations in the training set. For the penguins data:

```{r}
#| label: ex03_sqrt
sqrt(nrow(peng_train))
```

To fit the models, run:

```{r}
#| label: ex03_fit
models = list(
  # k-nearest neighbors
  knn = nearest_neighbor("classification", neighbors = 15),
  # Classification tree
  tree = decision_tree("classification"),
  # Multinomial regression (a linear model)
  multinomial = multinom_reg("classification"),
  # Support vector machine
  svm = svm_rbf("classification")
)

form =
  species ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g

sapply(models, function(model) {
  fitted = fit(model, form, peng_train)

  predicted = predict(fitted, peng_test)$.pred_class
  sum(peng_test$species == predicted) / nrow(peng_test)
})
```

All of the models predict with accuracy above 92%, and three are within 1 of
97%. This example shows that for this data set, the choice of model has a
relatively minor effect on accuracy. That's the case for many data sets. Having
enough observations and selecting or engineering relevant features is far more
important.


Selecting Features
------------------

![In a typical data science workflow, modeling is just one step among many and
depends on the other steps. As with writing essays, modeling is usually a
process of iterative refinement.](../figures/data_science_workflow.png)

In practice, the choice of method has less impact on predictive performance
than the features and number of observations in the training data. Different
methods *do make different assumptions*, and it's important to think through
whether the assumptions are satisfied, but no method is a substitute for data.

Some models, such as random forest models, can automatically select the most
relevant features for prediction in a data set. With generalized linear models,
you can use **regularization** methods, such as LASSO or Elastic Net, which add
a penalty to the model's loss function to encourage zeroing coefficients for
unimportant features. These techniques can be helpful for screening features
even if you ultimately plan to use a different model. You can also use
exploratory analysis to assess which features have predictive power, and should
use any expert knowledge you have about the data or population from which it
was sampled.

Transforming features to create new features, a process known as **feature
engineering**, can also improve model performance. Unfortunately, it can be
difficult to determine which features to engineer unless you have deep
knowledge about the data and problem you're modeling.

::: {.callout-tip}
[Feature Engineering and Selection: A Practical Approach for Predictive
Models][fes] by M. Kuhn & K. Johnson is a clear and thorough reference.
:::



Evaluating Models {#sec-evaluating-models}
-----------------

Perhaps the most informative classification model performance statistic is a
**confusion matrix**, a cross-tabulation of actual and predicted classes.
Unlike an accuracy estimate, a confusion matrix shows how well the model
performs for each class. You can also use the confusion matrix to compute a
variety of other statistics:

![
  Statistics you can use to evaluate classifier performance when there are two
  classes. Most of these generalize to more classes. Visualization inspired by
  [Wikipedia's diagnostic testing diagram][wiki-perf].
](../figures/performance_metrics.png)

[wiki-perf]: https://en.wikipedia.org/wiki/Template:Diagnostic_testing_diagram

Four widely-used statistics are:

* **Precision**, which is the proportion of predictions for a class that are
  correct. High precision for a given class indicates that a model is generally
  correct when it predicts the class, but might fail to recognize some members
  of the class.
* **Recall** or **sensitivity**, which is the proportion of a class' members
  that are correctly predicted. High recall for a given class indicates the
  model correctly recognizes members of the class, but might also predict
  members of other classes are in the class.
* **Specificity**, which is recall for the second class in a two-class problem.
* The **F1 score**, which is the harmonic mean of precision and recall. As a
  harmonic mean, the F1 score is close to 0 when precision *or* recall are and
  close to 1 when both precision *and* recall are.

You learn more about precision, recall, and the F1 score from [this interactive
guide][mlu-pr].

[mlu-pr]: https://mlu-explain.github.io/precision-recall/

The figure shows the situation for a two-class problem. You can also compute
all of these statistics on a per-class basis when there are more than two
classes (treat all other classes as the negative class). Moreover, you can
compute overall precision, recall, and F1 score by averaging the class
statistics. This is known as **macro-averaging**.

Another widely-used measure of performance for classification models is the
**receiver operating characteristic** (ROC) curve and accompanying **area under
the curve** (AUC). These measures are natural for classifiers that predict the
probabilty of each class and use the predicted probabilities with a threshold
decision rule to select the predicted class, because they show the behavior of
the model for different choices of threshold. For models that do not predict
class probabilities, the ROC curve does not exist, although researchers have
devised ways to generalize the ROC curve to some of these models. You can learn
more about ROC curves and AUC from [this interactive guide][mlu-roc].

[mlu-roc]: https://mlu-explain.github.io/roc-auc/


### Partitioning Data

In @sec-four-models, we split the Palmer penguins data set into two parts: a
**training set** to use for training models and a **testing set** to use for
testing models. Splitting data sets this way ensures that performance
statistics computed on the testing set are accurate. If you compute performance
statistics for a model with the same data set you used for training, the
statisitics will have an optimistic bias. In other words, accuracy, precision,
recall and other positive statistics will be overestimated, while error rate
and other negative statistics will be underestimated.

It's important that the testing set is not used to train models at any point in
the modeling process. Even using the testing set to select models or
hyperparameters can bias the testing set estimates.

Since performance statistics are also helpful for selecting models and
hyperparameters, and the training set will produce biased estimates, it's often
useful to split off a small part of the training set as a **validation set**.
The validation set should only be used for selecting models and
hyperparameters, not for training or final performance statistics. You can
learn more about training, validation, and testing sets from [this interactive
guide][mlu-train-test].

[mlu-train-test]: https://mlu-explain.github.io/train-test-validation/

You might wonder whether it's inefficient to split the data set into so many
subsets, since models generally perform better when they are trained on larger
data sets. You'd be right to wonder---using a fixed training set and validation
set can create a pessimistic bias in the performance statistics computed on the
validation set. It also makes the statistics sensitive to any outliers or
anomalous observations that end up in the validation set.

**$k$-fold cross validition** is one way to address the problems with using a
fixed training set and validation set. In a $k$-fold cross-validation scheme,
the training set is split into $k$ random, equal-size subsets called **folds**.
Then the model is fitted $k$ times, giving each fold a chance to be the
validation set (the remaining folds are used as a training set). This produces
$k$ estimates of any performance statistics of interest, which can be averaged
to get an overall estimate. Cross-validation uses more of the data to train the
model, which makes the performance statistic estimates more accurate. By
averaging the estimates over several folds, it also ensures that the overall
estimate is not too sensitive to unusual observations in a single fold. You can
learn more about cross-validation from [this interactive guide][mlu-cv].

[mlu-cv]: https://mlu-explain.github.io/cross-validation/

The main drawback of cross-validition is that the model must be trained more
than once. For some models, training is computationally intensive, and training
multiple times can be prohibitively so. For this reason, it's common to set $k
= 5$ or $k = 10$. Smaller values of $k$ produce noisier estimates. An extreme
case, known as **leave-one-out cross-validation** (LOOCV), sets $k$ to the size
of the training set, so that each observation is a fold. LOOCV is completely
deterministic, whereas for any other $k$ the cross-validation estimates vary
depending on which observations end up in each fold.


### Example: Cross-Validated kNN

Let's revisit the Palmer penguins example that begin in @sec-penguins one more
time. In the original example, we trained a $k$-nearest neighbors model and
arbitrarily selected $k = 10$. In @sec-four-models, we selected $k = 15$
instead based on a rule of thumb. In practice, the best way to select $k$ is by
estimating performance for several different values of $k$. We can use cross
validation to compute the estimates.

The rsample package provides a function `vfold_cv` to compute cross-validation
folds. The function returns a data frame with the folds stored in the `splits`
column. For each fold, you can use the `training` function to get the training
set and the `testing` function to get the validation set.

For this example, let's use 5-fold cross-validation to keep the computation
time reasonable. Using more folds would give us more accurate estimates. We'll
try all odd values of $k$ from 1 to approximately the minimum number of
observations in a training set:

```{r}
#| label: ex04_folds
# Set a seed to make this reproducible.
set.seed(1010)
# 5-fold cross-validation
folds = vfold_cv(peng_train, 5)

# Get a rough estimate of the minimum training set size.
n_train = floor((4 / 5) * 0.75 * nrow(peng_train))
# Try k = 1, 3, 5, ..., n_train.
ks = seq(1, n_train, 2)

form =
  species ~ bill_length_mm + bill_depth_mm + flipper_length_mm + body_mass_g
```

:::{.callout-caution}
Using $k$-fold cross-validation this way can produce inaccurate estimates when
there are more observations for some classes than others (as is the case here).

It would be better to use **stratified $k$-fold cross-validation**, which
attempts to preserve the overall class proportions within each fold. You can do
this with the `vfold_cv` function by setting the `strata` parameter to the
feature you want to use for stratification (in this case, we would use
`species`).
:::

Now we can use nested apply functions or loops to iterate over the $k$ and
folds. For each fold, we'll get the training set and train the model with the
current $k$, then predict on the test set. We can use the predictions to
compute performance statistics. To keep the example simple, let's compute
overall accuracy, although in practice it's important to look at other
statistics, including per-class statistics. The code to carry out these steps
is:

```{r}
#| label: ex04_cv
# Loop over k values.
error_rates = sapply(ks, function(k) {
  # Loop over folds.
  sapply(folds$splits, function(fold) {
    # Fit the model on the training set.
    train = training(fold)
    model = nearest_neighbor("classification", neighbors = k,
      weight_func = "rectangular")
    fitted = fit(model, form, train)

    # Predict on the test set.
    test = testing(fold)
    preds = predict(fitted, test)
    sum(test$species != preds$.pred_class) / nrow(test)
  })
})
```

The result from the code is a matrix of estimates. Each row corresponds to one
value of $k$, while each column corresponds to one fold. Let's compute an
overall estimate for each $k$ with the `colMeans` function, and then plot the
estimates against $k$.

```{r}
#| label: ex04_plot
estimates = data.frame(
  k = ks,
  error = colMeans(error_rates)
)

ggplot(estimates) + aes(x = k, y = error) + geom_line()
```

For this particular data set, it looks like $k = 1$ provides the best accuracy.
This may seem somewhat surprising because of the bias-variance tradeoff, and it
is a good idea to double-check your code in situations like this. However, for
this data set, each penguin species is so well-separated by the four features
that $k = 1$ really does work well. If we wanted the use the model for new
samples of penguins, it might still be a good idea to choose a slightly higher
$k$ value just in case this data set happens to have less variation than the
population.
